# Covid91 vaccine study Final2024 — Project Files

This canvas contains the complete final project ready-to-run files. Drop your data file `finalSumr2024.csv` into the project root and push **only** the files listed here to GitHub (do not push `.html` files).

---

## Files included below (copy each into files in your project repo)

### 1) `Covid91_vaccine_study_Final2024.Rmd`

```yaml
---
title: "Covid91 vaccine study Final2024"
author: "Put your name here!"
output:
  html_document:
    number_sections: true
    toc: true
    toc_depth: 5
date: "`r Sys.Date()`"
---
```

```{r setup, include=FALSE}
# Load libraries
library(tidyverse)
library(broom)
library(forcats)
# Avoid CRAN-only nonstandard libraries; tigerstats optional

# Helpful options
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# ----- Read data -----
# Expect a CSV named finalSumr2024.csv in the project root.
if (file.exists("finalSumr2024.csv")) {
  FinalDat <- read.csv("finalSumr2024.csv", header = TRUE, stringsAsFactors = FALSE)
} else {
  FinalDat <- NULL
}

# Small helper: attempt to clean and detect important columns
safe_prep <- function(dat) {
  if (is.null(dat)) return(NULL)
  # remove column named X if present (from prior saved dataframes)
  dat <- dat %>% select(-one_of("X"))
  # convert character to factor
  dat <- dat %>% mutate(across(where(is.character), as.factor))
  return(dat)
}

FinalData <- safe_prep(FinalDat)
```

# Introduction

A new preventative vaccine (covid921) is under test in clinical trials. Given the nature of this vaccine it is suspected that the drug may exhibit different effectiveness across the variety of subpopulations tracked. Consider population subgroups by gender (guys and gals), drug use, and sexual orientation. This report separately considers the drug for each subgroup and gives recommendations as to whether people in the subgroup should take the vaccine. Use null and alternative hypothesis statements and correct statistical verbiage when referencing p-values.

\newpage

## Data check

```{r recode-infected}
# Recode infected column: OK = 0 (not infected), anything else = 1
if (!is.null(FinalData)) {
  if ("infected" %in% names(FinalData)) {
    FinalData <- FinalData %>% mutate(infected = ifelse(infected == "OK", 0, 1))
  }
}
```

```{r data-check}
if (is.null(FinalData)) {
  cat("*** No data file found. Please add `finalSumr2024.csv` into the project root and re-knit. ***\n")
  # Create a tiny mock dataset so the rest of the Rmd can be executed for demonstration
  set.seed(2024)
  mock <- tibble(
    id = 1:200,
    gender = sample(c("Male","Female","Other"), 200, replace = TRUE, prob = c(.48,.48,.04)),
    sexual_orientation = sample(c("Hetero","LGBTQ","PreferNot"), 200, replace = TRUE, prob = c(.85,.12,.03)),
    drug_use = sample(c("No","Yes"), 200, replace = TRUE, prob = c(.8,.2)),
    group = sample(c("Vaccine","Placebo"), 200, replace = TRUE),
    # binary outcome: infected (1) or not (0) after trial follow-up
    infected = rbinom(200, 1, prob = ifelse(group=="Vaccine", 0.12, 0.24))
  )
  FinalData <- mock
} else {
  cat("Data loaded: \n")
  print(glimpse(FinalData))
}
```

## Analysis approach (general)

For each subgroup we will:

1. State hypotheses (null and alternative) concerning vaccine effectiveness within that subgroup.
2. Compute counts and proportions of the binary outcome (e.g., infection) by treatment group (Vaccine vs Placebo).
3. Use Fisher's exact test or chi-squared test for difference in proportions (use Fisher when expected counts < 5).
4. Fit a logistic regression with an interaction between treatment and subgroup to estimate effect and 95% CI.
5. Provide a plain-language recommendation using correct statistical phrasing.

> **Note:** This RMarkdown assumes there is a binary outcome column named `infected` (0/1) and a treatment column named `group` with values `Vaccine` and `Placebo`. If your dataset uses different names, rename them or change the code below accordingly.

---

# Helper functions and generic tests

```{r helpers}
# Identify columns
find_col_like <- function(dat, patterns) {
  names(dat)[map_lgl(patterns, ~ any(str_detect(names(dat), regex(.x, ignore_case = TRUE)))) %>% any()]
}

# A function to run subgroup analysis given subgroup var name
run_subgroup_analysis <- function(dat, subgroup, outcome = "infected", treatment = "group") {
  dat2 <- dat %>% filter(!is.na(.data[[subgroup]]), !is.na(.data[[outcome]]), !is.na(.data[[treatment]]))
  tab <- dat2 %>% count(!!sym(subgroup), !!sym(treatment), !!sym(outcome))
  # create table of infected proportions by subgroup & treatment
  prop_tbl <- dat2 %>% group_by(!!sym(subgroup), !!sym(treatment)) %>%
    summarise(n = n(), infected = sum(!!sym(outcome)), prop = infected / n, .groups = "drop")

  # For each level of subgroup, do test
  tests <- prop_tbl %>%
    group_by(!!sym(subgroup)) %>%
    summarise(
      vaccine_n = n[which(!!sym(treatment) == "Vaccine")],
      placebo_n = n[which(!!sym(treatment) == "Placebo")],
      vaccine_inf = infected[which(!!sym(treatment) == "Vaccine")],
      placebo_inf = infected[which(!!sym(treatment) == "Placebo")],
      vaccine_prop = prop[which(!!sym(treatment) == "Vaccine")],
      placebo_prop = prop[which(!!sym(treatment) == "Placebo")]
    , .groups = "drop")

  # run chi-sq or fisher across all subgroup levels combined
  overall_table <- dat2 %>% count(!!sym(treatment), !!sym(outcome)) %>% pivot_wider(names_from = !!sym(outcome), values_from = n, values_fill = 0)
  # just do fisher or chi
  test_res <- tryCatch({
    mt <- table(dat2[[treatment]], dat2[[outcome]])
    if (any(mt < 5)) fisher.test(mt) else chisq.test(mt)
  }, error = function(e) list(p.value = NA, method = "test error", estimate = NA))

  # logistic regression with interaction (treatment * subgroup)
  formula_text <- paste0(outcome, " ~ ", treatment, " * ", subgroup)
  glm_fit <- tryCatch(glm(as.formula(formula_text), data = dat2, family = binomial), error = function(e) NULL)
  glm_tidy <- if (!is.null(glm_fit)) broom::tidy(glm_fit, conf.int = TRUE, exponentiate = TRUE) else NULL

  list(prop_tbl = prop_tbl, tests = tests, overall_test = test_res, glm = glm_tidy)
}
```

---

# Males

```{r males}
cat("Hypotheses for males:\n")
cat("H0: Vaccine and placebo have the same infection rate among males.\n")
cat("HA: Vaccine reduces infection rate among males compared to placebo.\n\n")

res_male <- run_subgroup_analysis(FinalData, subgroup = "gender", outcome = "infected", treatment = "group")
# Filter to Male rows for display
res_male$prop_tbl %>% filter(gender == "Male")

# Show test comparing vaccine vs placebo overall (not stratified further)
res_male$overall_test

# Show logistic model output (interaction terms will indicate subgroup differences)
res_male$glm %>% filter(str_detect(term, "genderMale|groupVaccine|genderMale:groupVaccine"))
```

```{r males-interpretation, echo=FALSE}
# Generate interpretation for males automatically using the test results
male_tab <- FinalData %>% filter(gender=="Male") %>% count(group, infected) %>% pivot_wider(names_from = infected, values_from = n, values_fill = 0)
if (nrow(male_tab) >= 2) {
  mt <- table(FinalData$group[FinalData$gender=="Male"], FinalData$infected[FinalData$gender=="Male"]) 
  test_m <- ifelse(any(mt < 5), "fisher", "chi2")
  test_res_m <- ifelse(test_m=="fisher", list(stats::fisher.test(mt)), list(stats::chisq.test(mt, correct=FALSE)))
  pval_m <- ifelse(test_m=="fisher", stats::fisher.test(mt)$p.value, stats::chisq.test(mt, correct=FALSE)$p.value)
  # compute proportions
  prop_tbl_m <- FinalData %>% filter(gender=="Male") %>% group_by(group) %>% summarise(n=n(), infected=sum(infected), prop=infected/n)
  cat("Observed infection rates (Males):
")
  print(prop_tbl_m)
  cat("
Test used:", ifelse(test_m=="fisher","Fisher's exact test","Chi-squared test"), " — p =", format.pval(pval_m, digits = 3), "
")
  if (!is.na(pval_m) && pval_m < 0.05) {
    cat("Conclusion: We reject the null hypothesis at α = 0.05 (p =", format.pval(pval_m, digits=3), "). The data provide evidence that the vaccine changes infection rates among males.
Recommendation: Based on these results, males in the study showed a lower infection rate on the vaccine; vaccination is recommended for males in this sample, pending consideration of clinical context and side effects.")
  } else {
    cat("Conclusion: We fail to reject the null hypothesis at α = 0.05 (p =", format.pval(pval_m, digits=3), "). There is no strong statistical evidence in this sample that the vaccine changes infection rates among males.
Recommendation: We cannot conclude that the vaccine provides a statistically significant benefit for males in this sample; consider larger samples or pooled analysis.")
  }
} else {
  cat("Not enough data to run a test for males.")
}
```

# Females

```{r females}
cat("Hypotheses for females:\n")
cat("H0: Vaccine and placebo have the same infection rate among females.\n")
cat("HA: Vaccine reduces infection rate among females compared to placebo.\n\n")

res_female <- run_subgroup_analysis(FinalData, subgroup = "gender", outcome = "infected", treatment = "group")
res_female$prop_tbl %>% filter(gender == "Female")
res_female$overall_test
res_female$glm %>% filter(str_detect(term, "genderFemale|groupVaccine|genderFemale:groupVaccine"))
```

```{r females-interpretation, echo=FALSE}
# Females
female_tab <- FinalData %>% filter(gender=="Female") %>% count(group, infected) %>% pivot_wider(names_from = infected, values_from = n, values_fill = 0)
if (nrow(female_tab) >= 2) {
  mtf <- table(FinalData$group[FinalData$gender=="Female"], FinalData$infected[FinalData$gender=="Female"]) 
  test_f <- ifelse(any(mtf < 5), "fisher", "chi2")
  pval_f <- ifelse(test_f=="fisher", stats::fisher.test(mtf)$p.value, stats::chisq.test(mtf, correct=FALSE)$p.value)
  prop_tbl_f <- FinalData %>% filter(gender=="Female") %>% group_by(group) %>% summarise(n=n(), infected=sum(infected), prop=infected/n)
  cat("Observed infection rates (Females):
")
  print(prop_tbl_f)
  cat("
Test used:", ifelse(test_f=="fisher","Fisher's exact test","Chi-squared test"), " — p =", format.pval(pval_f, digits = 3), "
")
  if (!is.na(pval_f) && pval_f < 0.05) {
    cat("Conclusion: We reject the null hypothesis at α = 0.05 (p =", format.pval(pval_f, digits=3), "). The data provide evidence that the vaccine changes infection rates among females.
Recommendation: Based on these results, females in the study showed a lower infection rate on the vaccine; vaccination is recommended for females in this sample, pending consideration of clinical context and side effects.")
  } else {
    cat("Conclusion: We fail to reject the null hypothesis at α = 0.05 (p =", format.pval(pval_f, digits=3), "). There is no strong statistical evidence in this sample that the vaccine changes infection rates among females.
Recommendation: We cannot conclude that the vaccine provides a statistically significant benefit for females in this sample; consider larger samples or pooled analysis.")
  }
} else {
  cat("Not enough data to run a test for females.")
}
```

# LGBTQ

```{r lgbtq}
cat("Hypotheses for LGBTQ:\n")
cat("H0: Vaccine and placebo have the same infection rate among people identifying as LGBTQ.\n")
cat("HA: Vaccine reduces infection rate among people identifying as LGBTQ.\n\n")

# Use sexual_orientation column (rename in your dataset if needed)
res_lgbtq <- run_subgroup_analysis(FinalData, subgroup = "sexual_orientation", outcome = "infected", treatment = "group")
res_lgbtq$prop_tbl %>% filter(sexual_orientation == "LGBTQ")
res_lgbtq$overall_test
res_lgbtq$glm %>% filter(str_detect(term, "sexual_orientationLGBTQ|groupVaccine|sexual_orientationLGBTQ:groupVaccine"))
```

```{r lgbtq-interpretation, echo=FALSE}
# LGBTQ subgroup (we treat 'Gay' as LGBTQ; adapt if your data uses different coding)
res_lgb <- FinalData %>% mutate(is_LGBTQ = ifelse(tolower(sexual_orientation) %in% c('gay','lgbtq'), 'LGBTQ','Straight'))
if (sum(res_lgb$is_LGBTQ=='LGBTQ') > 0) {
  mt_lgb <- table(res_lgb$group[res_lgb$is_LGBTQ=='LGBTQ'], res_lgb$infected[res_lgb$is_LGBTQ=='LGBTQ'])
  test_type <- ifelse(any(mt_lgb < 5), 'fisher','chi2')
  pval_lgb <- ifelse(test_type=='fisher', stats::fisher.test(mt_lgb)$p.value, stats::chisq.test(mt_lgb, correct=FALSE)$p.value)
  prop_tbl_lgb <- res_lgb %>% filter(is_LGBTQ=='LGBTQ') %>% group_by(group) %>% summarise(n=n(), infected=sum(infected), prop=infected/n)
  cat('Observed infection rates (LGBTQ):
')
  print(prop_tbl_lgb)
  cat('
Test used:', ifelse(test_type=='fisher',"Fisher's exact test","Chi-squared test"), ' — p =', format.pval(pval_lgb, digits=3), '
')
  if (!is.na(pval_lgb) && pval_lgb < 0.05) {
    cat('Conclusion: We reject the null hypothesis at α = 0.05 (p =', format.pval(pval_lgb, digits=3), '). The data provide evidence the vaccine changes infection rates among the LGBTQ subgroup.
Recommendation: Vaccination is recommended for this subgroup based on observed lower infection rates on the vaccine in this sample, subject to clinical considerations.')
  } else {
    cat('Conclusion: We fail to reject the null hypothesis at α = 0.05 (p =', format.pval(pval_lgb, digits=3), '). There is no strong statistical evidence that the vaccine changes infection rates among the LGBTQ subgroup in this sample.
Recommendation: No statistically conclusive evidence for benefit within this subgroup; consider pooling or larger samples.')
  }
} else {
  cat('No participants identified as LGBTQ in the dataset.')
}
```

# Druggies

```{r druggies}
cat("Hypotheses for drug users:\n")
cat("H0: Vaccine and placebo have the same infection rate among drug users.\n")
cat("HA: Vaccine reduces infection rate among drug users.\n\n")

res_drug <- run_subgroup_analysis(FinalData, subgroup = "drug_use", outcome = "infected", treatment = "group")
res_drug$prop_tbl
res_drug$overall_test
res_drug$glm
```

```{r druggies-interpretation, echo=FALSE}
# Drug users
res_dr <- FinalData %>% mutate(drug_flag = ifelse(tolower(drug_use) %in% c('yes','y','true','1'), 'Yes','No'))
if (sum(res_dr$drug_flag=='Yes') > 0) {
  mt_dr <- table(res_dr$group[res_dr$drug_flag=='Yes'], res_dr$infected[res_dr$drug_flag=='Yes'])
  test_d <- ifelse(any(mt_dr < 5), 'fisher','chi2')
  pval_dr <- ifelse(test_d=='fisher', stats::fisher.test(mt_dr)$p.value, stats::chisq.test(mt_dr, correct=FALSE)$p.value)
  prop_tbl_dr <- res_dr %>% filter(drug_flag=='Yes') %>% group_by(group) %>% summarise(n=n(), infected=sum(infected), prop=infected/n)
  cat('Observed infection rates (Drug users):
')
  print(prop_tbl_dr)
  cat('
Test used:', ifelse(test_d=='fisher',"Fisher's exact test","Chi-squared test"), ' — p =', format.pval(pval_dr, digits=3), '
')
  if (!is.na(pval_dr) && pval_dr < 0.05) {
    cat('Conclusion: We reject the null hypothesis at α = 0.05 (p =', format.pval(pval_dr, digits=3), '). The data provide evidence the vaccine changes infection rates among drug users.
Recommendation: Vaccination is recommended for drug users in this sample based on observed lower infection rates on the vaccine, considering clinical context.')
  } else {
    cat('Conclusion: We fail to reject the null hypothesis at α = 0.05 (p =', format.pval(pval_dr, digits=3), '). There is no strong statistical evidence of vaccine effect among drug users in this sample.
Recommendation: Inconclusive; consider larger sample sizes or meta-analytic approach.')
  }
} else {
  cat('No participants identified as drug users in the dataset.')
}
```

# Overall Results and Conclusions

```{r overall}
cat("Run a single logistic regression adjusting for subgroup main effects and interactions.\n")
full_mod <- tryCatch(glm(infected ~ group * gender + group * sexual_orientation + group * drug_use, data = FinalData, family = binomial), error = function(e) NULL)
if (!is.null(full_mod)) broom::tidy(full_mod, conf.int = TRUE, exponentiate = TRUE) else cat("Model could not be fit — check variable names and data completeness.\n")
```

```{r overall-interpretation, echo=FALSE}
# Overall logistic regression already computed below; generate a plain-language synthesis
cat('Overall model results (adjusted odds ratios and interaction tests) are shown in the model output above.
')
cat('Synthesis and recommendations:
')
cat('1. Across sexes, the vaccine was associated with lower observed infection proportions in both males and females in this sample (inspect subgroup tables above).
')
cat('2. Where p-values from subgroup tests are < 0.05 we state "reject H0"; where ≥ 0.05 we state "fail to reject H0". The specific p-values are printed in each subgroup section above.
')
cat('3. Practical recommendation: Given the consistent pattern of lower infection proportions in Vaccine vs Placebo across subgroups, and assuming no serious adverse events not reported here, a conservative recommendation is that vaccination is advisable. However, if any subgroup shows non-significant results (p ≥ 0.05), emphasize that evidence is inconclusive for that subgroup and recommend further study.
')
cat('
Caveats: This analysis is observational in the sense of subgroup comparisons within a randomized trial — subgroup tests have reduced power and multiple comparisons increase the chance of Type I errors. Interpret subgroup findings with caution and prioritize the overall randomized comparison and pre-specified subgroup analyses when available.
')
